---
title: summary of maxent modeling 
author: Daniel Carver
output:
  html_document:
    code_folding: hide
    highlight: tango
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
      smooth_scroll: no
---


```{r echo=FALSE, message=FALSE, warning=FALSE}

library(knitr)
library(markdown)
library(rmarkdown)
library("tmap")
library("raster")
library("DT")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# function for flitering list based on character values 
include <- function (theList, toMatch){
  matches <- unique (grep(paste(toMatch,collapse="|"), 
                          theList, value=TRUE))
  return(matches)
}
# taxa <- "Cucurbita_pepo_subsp._ovifera_var._ozarkana"
# run<- "biome"
```

## Summary of the models for `r taxa` with a native area boundary of `r run`. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# taxa is the 
baseDir <- paste0(base_dr, '/', taxa,'/',run)
csv <- list.files(baseDir, pattern = ".csv", recursive = TRUE , full.names = TRUE)

```

### Evaluation metrics 

The modeling process was ran 5 time to attempt to account for random variation that is a potential with this type of model. The table below show that statistical results of the 5 runs. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# evalMetric <- include(csvs,"eval_metrics_rep.csv")
DT::datatable(read.csv(include(csv,"eval_metrics_rep.csv")))
```

For summary the medain run results is defined here. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# evalMetric <- include(csvs,"eval_metrics_rep.csv")
DT::datatable(read.csv(include(csv,"eval_metrics.csv")))
```

### Conservation 

The table below show that exsitu conservation summary 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# exSummary <- include(csv, "exsitu/summary.csv")
DT::datatable(read.csv(include(csv, "exsitu/summary.csv")))
```


The table below show that insitu conservation summary 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# inSummary <- include(csv, "insitu/summary.csv")
DT::datatable(read.csv(include(csv, "insitu/summary.csv")))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
tif <- list.files(path = baseDir, pattern = '.tif', recursive = TRUE, full.names = TRUE)
median <- raster(include(tif, "spdist_median"))
thrshold <- raster(include(tif, "spdist_thrsld"))
sd <- raster(include(tif, "spdist_sd"))
tmap_mode("view")

```
### Maps of model outputs. 

Map of the median results from maxent models

```{r echo=FALSE, message=FALSE, warning=FALSE}
qtm(shp = median)
```

Map of the standard devation results from maxent models
```{r echo=FALSE, message=FALSE, warning=FALSE}
qtm(shp = sd)
```

Map of the threshold results from maxent models
```{r echo=FALSE, message=FALSE, warning=FALSE}
qtm(shp = thrshold)
```
