---
title: Summary of Cucurbita Habitat Modeling 
author: Daniel Carver. Colin Khoury 
output:
  html_document:
    code_folding: hide
    highlight: tango
    theme: yeti
    toc: yes
---


```{r echo=FALSE, message=FALSE, warning=FALSE}
# install.packages("tmap")
# install.packages("sf")
# install.packages("DT")
# install.packages("XML")
# install.packages("scales")
library(knitr)
library(markdown)
library(rmarkdown)
library("tmap")
library("raster")
library("DT")
# taxa <- "Cucurbita_cordata"

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# function for flitering list based on character values 
include <- function (theList, toMatch){
  matches <- unique (grep(paste(toMatch,collapse="|"), 
                          theList, value=TRUE))
  return(matches)
}
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# taxa is the 
###DC
baseDir <- paste0(base_dr, '/', taxa)
csv <- list.files(baseDir, pattern = ".csv", recursive = TRUE , full.names = TRUE)
```


### Summary Statistics for Model Runs for `r taxa`

This table show the statistical results of all median models. 
```{r echo=FALSE, message=FALSE, warning=FALSE}

dfSum <- data.frame()
runs <- list.dirs(baseDir,recursive = FALSE , full.names = FALSE)
metrics <- include(csv, "eval_metrics.csv")
for(i in 1:length(runs)){
  value <- grepl(runs[i], metrics, fixed = TRUE)
  if(any(value)){
  dt1 <- read.csv(metrics[value])
  dt1$run <- runs[i]
  dfSum <- rbind(dfSum, dt1)
  }else{
    next
  }
}

DT::datatable(dfSum)
```

### All Models Addative Map
This map combines all threshold maps into one. The darkest areas are where all models agree that the location represents potential suitable habitat.  
```{r echo=FALSE, message=FALSE, warning=FALSE}
tif <- list.files(path = baseDir, pattern = '.tif', recursive = TRUE, full.names = TRUE)
thres <- include(tif, "spdist_thrsld")
rasters <- lapply(thres,raster)
size <- lapply(rasters, ncell)
maxRas <- which.max(size)

ras1 <- rasters[maxRas]
ras1 <- ras1[[1]]
rastersThin <- rasters[-c(maxRas)]

ras1[is.na(ras1[])] <- 0 


#build the first raster here 
if(length(thres) > 0 ){
  for(i in 1:length(rastersThin))
    {
    rast2 <- grepl(runs[i], metrics, fixed = TRUE)
    if(any(rast2))
      {
      ras <- raster(thres[rast2])
      ras[is.na(ras[])] <- 0 
      ras <- extend(ras, extent(ras1), value = 0)
      ras1 <- ras1 + ras
      }else
    {
    next
    }
  }
}else{
  ras1 <- NA
}

ras1[ras1[] <= 0] <- NA 




tmap_mode("view")
if(class(ras1) == "RasterLayer"){
  qtm(shp = ras1)
}else{
  print("no rasters are present")
}
```
This map above is composed of `r length(rasters)` model results. 


### Biome and Ecoregion Maps

The map below shows the combination map of our two most conservative models. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
for(class in rasters){
  bio <- grepl("bio", metrics, fixed = TRUE)
  bioRast <- rasters[bio][[1]]
}

for(class in rasters){
  eco <- grepl("ecoR", metrics, fixed = TRUE)
  ecoRast <- rasters[eco][[1]]
}

for(class in rasters){
  ecoV <- grepl("ecoV", metrics, fixed = TRUE)
  ecoRast1 <- rasters[ecoV][[1]]
}


bio1 <- bioRast
eco1 <- ecoRast
ecoV <- ecoRast1

ecoFull <- extend(eco1, extent(bio1), value = 0)
ecoFullv <- extend(ecoV, extent(bio1), value = 0)

eco1[is.na(eco1[])] <- 0 
bio1[is.na(bio1[])] <- 0 
ecoV[is.na(ecoV[])] <- 0 
ecoBio <- ecoFull + bioRast + ecoFullv


if(class(ecoBio) == "RasterLayer"){
  qtm(shp = ecoBio)
}else{
  print("no rasters are present")
}
```

### Ecoregion Variable Selection 5 reps 
```{r echo=FALSE, message=FALSE, warning=FALSE}
for(class in rasters){
  testr <- grepl("testR", metrics, fixed = TRUE)
    if(any(testr)){
    testr <- rasters[testr][[1]]
  }else{
  print("no rasters are present")
  }
}


if(class(testr) == "RasterLayer"){
  qtm(shp = testr)
}else{
  print("no rasters are present")
}
```





### EcoRegion Variable Selection 10 reps 
The map below is an example of the prediction map that is limited to ecoregions in which known presense locations are found. A varible selection process was perform before the model was ran. 
```{r echo=FALSE, message=FALSE, warning=FALSE}
if(class(ecoRast1) == "RasterLayer"){
  qtm(shp = ecoRast1)
}else{
  print("no rasters are present")
}

```

### EcoRegion 
The map below is an example of the prediction map that is limited to ecoregions in which known presense locations are found. 
```{r echo=FALSE, message=FALSE, warning=FALSE}
if(class(ecoRast) == "RasterLayer"){
  qtm(shp = ecoRast)
}else{
  print("no rasters are present")
}

```


### Biome 
The map below is an example of the prediction map that is limited to biomes in which known presense locations are found. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
if(class(bioRast) == "RasterLayer"){
  qtm(shp = bioRast)
}else{
  print("no rasters are present")
}
```

### Country

The map below is an example of the prediction map that is limited to the countries in which known presense locations are found. 
```{r echo=FALSE, message=FALSE, warning=FALSE}
for(class in rasters){
  country <- grepl("count", metrics, fixed = TRUE)
    if(any(country)){
    country <- rasters[country][[1]]
  }else{
  print("no rasters are present")
  }
}


if(class(country) == "RasterLayer"){
  qtm(shp = country)
}else{
  print("no rasters are present")
}
```


### World 

The map below is an example of the prediction map that is based on background points generated from around the world. The model is constained to the countries in which the species is native. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
for(class in rasters){
  v1 <- grepl('v1', metrics, fixed = TRUE)
  if(any(v1)){
    v1 <- rasters[v1][[1]]
  }else{
  print("no rasters are present")
  }
}




if(class(v1) == "RasterLayer"){
  qtm(shp = v1)
}else{
  print("no rasters are present")
}
```
